Inferring community characteristics in labelled networks

### 2021-05-23 nips/appendices.tex
10 pages

@@

\subsection{Algorithms}

@@

\clearpage
\subsection{Hyperparameter values}
\label{appdx:hyperparams}

>>>>

\subsection{SBM prior choice explanation}
\label{appdx:prior}

We have borrowed the priors proposed by \citet{Peixoto-Bayesian-Microcanonical} for $p(\psi | b)$. These, are repeated here for reference:
%
\begin{equation}
	p(\psi | b) = p(e | b) p(k | e, b) = \left[ \specialchoose{ \specialchoose{B}{2} }{ E} \right]^{-1} 
	\cdot \left[ \prod_r \frac{\prod_j \eta_j^r !}{n_r! q(e_r, n_r)} \right]
\end{equation}
%
Where $\specialchoose{n}{m}$ is shorthand for $\binom{n+m-1}{m} = \frac{(n+m-1)!}{(n-1)!(m)!}$ which can be thought of as the total number of distinct histograms with $n$ bins under the constraint they sum to $m$. $E = \frac{1}{2} \sum_{r,s} e_{rs}$ is the total number of edges in the graph. Importantly, $E$ is not allowed to vary and so $p(e|b)$ is uniform with respect to $e$. The variable $\eta_j^r$ is introduced to denote the number of vertices in block $r$ that have degree $j$. Formally, $\eta_j^r \coloneqq \sum_{i} \one\left\{b_i = r \right\} \one \left\{k_i = j \right\}$. Furthermore, $q(m, n)$ is the number of different histograms with at most $n$ non-zero bins that sum to $m$. $q(m, n)$ is related to but different from $\specialchoose{n}{m}$. Lastly, $e_r \coloneqq \sum_{s} e_{rs}$ is the total number of half edges in block r and $n_r \coloneqq \sum_{i} \one\{b_i = r\}$ is the number of vertices assigned to block $r$.

These were chosen carefully by \citet{Peixoto-Bayesian-Microcanonical} to more closely match the structure of empirical networks than simple uniform priors. We do not repeat his arguments here.

@@

\subsection{Algorithms}
\label{appdx:algorithms}

@@

\FloatBarrier
\subsection{Hyperparameter values}
\label{appdx:hyperparams}

### 2021-05-23 nips/experiments.tex
10 pages

	\item \textbf{Primary school dynamic contacts} \cite{schools} ($N=238, E=5539, D=13$) -- network of face-to-face contacts amongst students and teachers at a primary school in Lyon, France. Two nodes are connected if the two parties shared a face-to-face interaction over the course of the day. Vertex features include class membership, gender and whether or not the individual is a teacher or pupil. These data were collected on consecutive days in October 2009. We choose to analyse just the second day.

@@

We require metrics to assess the relative model evidence for the FFBM. This can be split into two separate components: the microcanonical SBM fit (concerned with the $b$-samples) and the fit of the feature-to-block generator (concerned with the $\theta$-samples). Starting with the SBM, $S(b)$ (equation \ref{eqn:dl-form}) can be interpreted as the description length of the partition imposed by $b$. In other words this is the amount of information needed to describe the graph and the associated SBM parameters. It is only natural to divide this quantity by the number of entities (nodes and edges) in our graph $N+E$ to allow for rough comparison between graphs. This defines a simple metric to gauge the fit of the SBM: the description length per entity averaged over the $b$-samples (equation \ref{eqn:mean-dl}):
%
\begin{equation}
	\bar{S}_e \coloneqq \frac{1}{(N+E) |\Tcal_b|} \sum_{t\in \Tcal_b} S \left( b^{(t)} \right)
	\label{eqn:mean-dl}
\end{equation}
%
However, to assess the performance of the feature-to-block predictor, we must partition the vertex set $[N]$ into training and test sets -- $\Gcal_0$ and $\Gcal_1$ respectively
\footnote{We choose to randomly create the partition on each experiment run such that a constant fraction $f$ of the available vertices go to form our training set $\Gcal$ and the remaining vertices are held out to form our test set $\Gcal^C$. For all the experiments $f=0.7$.}
. The $b$-chain is run using the whole network but we only use vertices $v \in \Gcal$ to train the $\theta$-chain. As $|\Gcal_0| \neq |\Gcal_1|$ sets, we cannot use the un-normalised log target $U$ (equation \ref{eqn:U-form}) for comparison as the total cross-entropy loss is scaled by the size of each set -- even though the prior term stays constant. We therefore must use the average cross-entropy loss over each set (equation \ref{eqn:cross-entropy-loss}).
%
\begin{equation}
	\bar{\Lcal}_\star \coloneqq \frac{1}{|\Tcal_\theta|} \sum_{t \in \Tcal_\theta} \Lcal_\star \left( \theta^{(t)} \right)
	\qquad \textrm{where} \qquad
	\Lcal_\star \left( \theta^{(t)} \right) \coloneqq \frac{1}{|\Gcal_\star|} \sum_{i \in \Gcal_\star}\sum_{j \in [B]} \hat{y}_{ij} \log \frac{1}{\phi_j \left(x_i; \theta^{(t)} \right)}
	\label{eqn:cross-entropy-loss}
\end{equation}
%
Where $\star \in \{0, 1\}$ has been introduced to toggle between training and test sets. Table \ref{tab:results} summarises the results for each experiment. We run each experiment $n=10$ times with the same hyper-parameter values\footnote{For a comprehensive list of the hyper-parameters used for each experiment please see appendix \ref{appdx:hyperparams}} and report the mean plus or minus one standard deviation.

However, we also wish to determine the power of the dimensionality reduction method. For this we apply equation \ref{eqn:c-star}, to reduce the dimension from $D$ to $D'$ with $k=1$. We then retrain the feature-block predictor using just the retained feature set $\Dcal'$ and report the loss over the training and test sets for the reduced classifier -- denoted $\bar{\Lcal}_0'$ and $\bar{\Lcal}_1'$ respectively. These values are also included in table \ref{tab:results} except for the polbooks dataset which is already very low dimension.

@@

This dataset was collected by \citet{polbooks}. We wish to determine whether political affiliation is a good predictor of the overall network structure. We choose to partition the network into $B=3$ communities as we only have this many distinct values for political affiliation (conservative, liberal or neutral). The inferred block memberships are given in figure \ref{fig:polbooks-graph}. We sample the block-generator parameters and plot the emprical mean and standard deviation of each sample on figure \ref{fig:polbooks-null}.

Indeed, from figure \ref{fig:polbooks-null}, we see that all 3 blocks have a distinct political affiliation as their largest positive component. This is strong evidence that political affiliation is indeed the axis which best predicts the 3-way natural partition of the graph into blocks. Nevertheless, for block index 1, although \verb*|neutral| is the best predictor of the ones we have available, the value is not particularly extreme. Perhaps, there is an unobserved feature that places a book into this block. However, by inspecting figure \ref{fig:polbooks-graph} we note that there are very few edges between blocks 2 and 3 (the detected conservative and liberal blocks as per figure \ref{fig:school-null}). Block 1 effectively acts as a separator between blocks 2 and 3; very few readers venture all the way across the political aisle for their next book. It is therefore, possible that the moderate value of the neutral weight for block 1 reflects the fact that this block contains a few peripheral conservative and liberal books at the corners of the block.

From table \ref{tab:results} we see that the training and test losses are very similar (the test loss mean is even below the training loss) and both are quite low in magnitude. This is very strong evidence that the features we have available (the author's political affiliation) are indeed the drivers of the high-level graphical structure at $B=3$.

@@

These data were originally collected by \citet{schools} to quantify the transmission opportunities for respiratory infections. However, we seek to ask which vertex features best describe how people interact with one another in a primary school context. The only vertex features we have available are school-class (one of 10 values - 2 per year group), gender and a distinction between teachers and pupils.

@@

As before, we proceed to sample the block-generator parameters $\theta$ and employ the dimensionality reduction heuristic with standard deviation multiplier $k=1$ to pick out the top $D'=10$ features. We then plot the weights for the surviving features $d \in \Dcal'$ on figure \ref{fig:school-null}. Immediately, we see that only the pupils' class memberships have survived (1A-5B); gender and teacher/student status have been discarded meaning that these are not good predictors of overall macro-structure.

@@

As for the training and test losses of the original and reduced classifiers. We see that the training loss stays roughly constant between the original and reduced classifiers $\bar{\Lcal}_0 \approx \bar{\Lcal}_0'$ but the test loss of the reduced classifier is ever-so-slightly decreased (within margin of error). This improved generalisation implies that the features we have discarded (gender and teacher/student status) do not impact the high-level graphical structure and simply lead to over-fitting in the original classifier. The features we have available do not capture the entire picture but they do quite well (low overall loss).

@@

We again apply the dimensionality reduction heuristic with $k=1$ to target $D'=10$ features in the reduced set. The parameter values for the surviving features are plotted on figure \ref{fig:fb-null}. The remaining features are those that best explain the high-level community structure. The majority of the surviving features are education related. More so than language, gender or surname - the school\footnote{This includes higher education institutes} you attend is the best explanatory variable for how the observed network's communities are partitioned at the highest level. Nevertheless, for $D'=10$ we do not have good explanations for the makeup of blocks 5-9 inclusive

Nevertheless, this example highlights a weakness of the method for high-dimensional feature spaces. When the feature dimension is very large, it becomes increasingly likely that a particular feature may uniquely identify a small set of nodes. If these nodes are all part of the same community then the classifier will overfit for that particular parameter. The regularisation term imposed by the prior goes some way to alleviating this problem but we see in figure \ref{fig:fb-null} that the feature \verb*|birthday-5| has a very high weight as it relates to block \textbf{XX}. It might be possible to shift the feature values such that they take values in $\{-1, 1\}$ rather than $\{0, 1\}$ but this approach falls down when input features are mutually correlated. Indeed, for features with discrete values such as the class memberships (1A-5B) of the school dataset it is preferred to keep $x_i \in \{0, 1\}^D$ as then each block can be readily accept more than one mutually exclusive feature-group (i.e. two school-classes are part of the same block).

The training and test losses (table \ref{tab:results}) are the highest of the considered networks. This suggests that the features we have available do not readily explain the high-level structure. Nonetheless, after reducing the problem dimension our test loss is not greatly increased which suggests that the features we do keep are the best from the ones we have available at predicting the overall network structure.

>>>>

	\item \textbf{Primary school dynamic contacts} \cite{schools} ($N=238, E=5539, D=13$) -- network of face-to-face contacts amongst students and teachers at a primary school in Lyon, France. Two nodes are connected if the two parties shared a face-to-face interaction over the course of the day. Vertex features include class membership (one of 10 values 1A-5B), gender and whether or not the individual is a teacher or pupil. These data were collected on consecutive days in October 2009. We choose to analyse just the second day.

@@

We require metrics to assess performance. This can be split into two separate components: the microcanonical SBM fit (concerned with the $b$-samples) and the fit of the feature-to-block generator (concerned with the $\theta$-samples). Starting with the SBM, $S(b)$ (equation \ref{eqn:dl-form}) can be interpreted as the description length of the partition imposed by $b$. It is only natural to divide this quantity by the number of entities (nodes and edges) in our graph $N+E$ to allow for rough comparison between graphs. This defines a simple metric to gauge the fit of the SBM: the description length per entity averaged over the $b$-samples (equation \ref{eqn:mean-dl}):
%
\begin{equation}
	\bar{S}_e \coloneqq \frac{1}{(N+E) |\Tcal_b|} \sum_{t\in \Tcal_b} S \left( b^{(t)} \right)
	\label{eqn:mean-dl}
\end{equation}
%
However, to assess the performance of the feature-to-block predictor, we must partition the vertex set $[N]$ into training and test sets: $\Gcal_0$ and $\Gcal_1$ respectively.\footnote{We choose to randomly create the partition on each experiment run such that a constant fraction $f$ of the available vertices go to form our training set $\Gcal_0$ and the remaining vertices are held out to form our test set $\Gcal_1$. For all the experiments $f=0.7$.}
The $b$-chain is run using the whole network but we only use vertices $v \in \Gcal_0$ to train the $\theta$-chain. As $|\Gcal_0| \neq |\Gcal_1|$ sets, we cannot use the un-normalised log target $U$ (equation \ref{eqn:U-form}) for comparison; the total cross-entropy loss is scaled by the size of each set but the prior term stays constant. We therefore must use the average cross-entropy loss over each set (equation \ref{eqn:cross-entropy-loss}):
%
\begin{equation}
	\bar{\Lcal}_\star \coloneqq \frac{1}{|\Tcal_\theta|} \sum_{t \in \Tcal_\theta} \Lcal_\star \left( \theta^{(t)} \right)
	\qquad \textrm{where} \qquad
	\Lcal_\star \left( \theta^{(t)} \right) \coloneqq \frac{1}{|\Gcal_\star|} \sum_{i \in \Gcal_\star}\sum_{j \in [B]} \hat{y}_{ij} \log \frac{1}{\phi_j \left(x_i; \theta^{(t)} \right)}
	\label{eqn:cross-entropy-loss}
\end{equation}
%
Where $\star \in \{0, 1\}$ has been introduced to toggle between training and test sets. Table \ref{tab:results} summarises the results for each experiment.\footnote{For a comprehensive list of the hyper-parameters used for each experiment please see appendix \ref{appdx:hyperparams}}

We also apply the dimensionality reduction method on the two higher dimensional datasets (the school and FB egonet). For this we leverage equation \ref{eqn:c-star}, to reduce the dimension from $D$ to $D'$ with $k=1$ to yield $c^*$. We then retrain the feature-block predictor using just the retained feature set $\Dcal'$ and report the loss over the training and test sets for the reduced classifier -- denoted $\bar{\Lcal}_0'$ and $\bar{\Lcal}_1'$ respectively. These values are also included in table \ref{tab:results}.

@@

We wish to determine whether political affiliation is a good predictor of the overall network structure. We choose to partition the network into $B=3$ communities as we only have this many distinct values for political affiliation (conservative, liberal or neutral).

From, figure \ref{fig:polbooks-null}, we see that all 3 blocks have a distinct political affiliation as their largest positive component. This is strong evidence that political affiliation is indeed the axis which best predicts the 3-way natural partition of the graph into blocks.

From figure \ref{fig:polbooks-graph}, we note that there are very few edges between blocks 2 and 3 (the detected conservative and liberal blocks as per figure \ref{fig:school-null}). Block 1 effectively acts as a separator between blocks 2 and 3; very few readers venture all the way across the political aisle for their next book.

From table \ref{tab:results} we see that the training and test losses are very similar (the test loss mean is even below the training loss) and both are low in magnitude. This is very strong evidence that the features we have available (the author's political affiliation) are indeed the drivers of the high-level graphical structure at $B=3$.

@@

@@

As before, we sample the block-generator parameters $\theta$ and employ the dimensionality reduction technique with standard deviation multiplier $k=1$ to pick out the top $D'=10$ features. We then plot the weights for the surviving features $d \in \Dcal'$ on figure \ref{fig:school-null}. Immediately, we see that only the pupils' class memberships have survived (1A-5B); gender and teacher/student status have been discarded meaning that these are not good predictors of overall macro-structure.

@@

As for the training and test losses of the original and reduced classifiers. We see that the training loss stays roughly constant between the original and reduced classifiers $\bar{\Lcal}_0 \approx \bar{\Lcal}_0'$ but the test loss of the reduced classifier is ever-so-slightly decreased (within margin of error). This improved generalisation implies that the features we have discarded (gender and teacher/student status) do not impact the high-level graphical structure and simply lead to over-fitting in the original classifier. The features we have available do not capture the entire picture but they do quite well (modestly low overall loss).

@@

We again apply the dimensionality reduction technique with $k=1$ to target $D'=10$ features in the reduced set. The remaining features (figure \ref{fig:fb-null}) are those that best explain the high-level community structure. The majority of the surviving features are education related. Nevertheless, for $D'=10$ we only have good explanations for the makeup of some of the detected blocks.

When the feature dimension is very large, it becomes increasingly likely that a particular feature may uniquely identify a small set of nodes. If these nodes are all part of the same community then the classifier will overfit for that particular parameter. The regularisation term imposed by the prior goes some way to alleviating this problem. Nevertheless, we see in figure \ref{fig:fb-null} that the feature \verb*|birthday-5| has a very high weight as it relates to block 1. It might be possible to shift the feature values such that they take values in $\{-1, 1\}$ rather than $\{0, 1\}$ but to accept more than one feature-group per block.

The training and test losses (table \ref{tab:results}) are the highest of the considered networks but nonetheless still low. This suggests that the features we have available only partly explain the high-level structure. Promisingly, after reducing the problem dimension our test loss barely changes, suggesting that the features we do keep are the best from the ones we have available at explaining the overall network structure.

### 2021-05-23 nips/ffbm.tex
10 pages

In our model, the block memberships vector $b$ is an intermediate latent variable and so we are not free to choose a prior for it. Nevertheless, as far as inference on the right-hand-side of figure \ref{fig:ffbm}, we regard $p(b | X)$ as a pseudo-prior on $b$. We can show (appendix \ref{appdx:b|x}) that our choice of prior for $p(\theta)$ in equation \ref{eqn:theta-prior} leads to a uniform $p(b | X)$ in equation \ref{eqn:b-pseudo-prior}.
%
\begin{equation}
	p(b | X) = \int p(b | X, \theta) p(\theta) d\theta = B^{-N}
	\label{eqn:b-pseudo-prior}
\end{equation}
%
This is a uniform distribution that only depends on the number of blocks $B$. This is an enormously important simplification as evaluating $p(b | X)$ does not require knowing the exact value $X$ takes and can be performed without an expensive Monte-Carlo integration over the $\theta$-domain. \citet{Peixoto-Bayesian-Microcanonical} proposes careful choices for the additional microcanonical SBM parameters $\psi$ which we adopt. Peixoto's idea is to write the joint prior on $(b, e, k)$ as a product of conditionals $p(b, e, k) = p(b) p(e | b) p(k | e, b)= p(b) p(\psi | b)$. For our purposes we must insert a conditioning on $X$, to form our pseudo-prior for $b$ and $\psi$, to give equation \ref{eqn:joint-pseudo-prior}.
%
\begin{equation}
	p(b, \psi | X) = p(b | X) p(\psi | b, X) = p(b | X) p(\psi | b)
	\label{eqn:joint-pseudo-prior}
\end{equation}
%
Where the simplification is made apparent by noting $(\psi \indep X) | b$. We then borrow the priors proposed by \citet{Peixoto-Bayesian-Microcanonical} for $p(\psi | b)$, repeated here for reference:
%
\begin{equation}
	p(\psi | b) = p(e | b) p(k | e, b) = \left[ \specialchoose{ \specialchoose{B}{2} }{ E} \right]^{-1} 
	\cdot \left[ \prod_r \frac{\prod_j \eta_j^r !}{n_r! q(e_r, n_r)} \right]
\end{equation}
%
Where $\specialchoose{n}{m}$ is shorthand for $\binom{n+m-1}{m} = \frac{(n+m-1)!}{(n-1)!(m)!}$ which can be thought of as the total number of distinct histograms with $n$ bins under the constraint they sum to $m$. $E = \frac{1}{2} \sum_{r,s} e_{rs}$ is the total number of edges in the graph. Importantly, $E$ is not allowed to vary and so $p(e|b)$ is uniform with respect to $e$. The variable $\eta_j^r$ is introduced to denote the number of vertices in block $r$ that have degree $j$. Formally, $\eta_j^r \coloneqq \sum_{i} \one\left\{b_i = r \right\} \one \left\{k_i = j \right\}$. Furthermore, $q(m, n)$ is the number of different histograms with at most $n$ non-zero bins that sum to $m$. $q(m, n)$ is related to but different from $\specialchoose{n}{m}$. Lastly, $e_r \coloneqq \sum_{s} e_{rs}$ is the total number of half edges in block r and $n_r \coloneqq \sum_{i} \one\{b_i = r\}$ is the number of vertices assigned to block $r$. Importantly, we have computable forms for $p(\theta)$ and $p(b, \psi | X)$ which will be useful for performing inference.

>>>>

In our model, the block memberships vector $b$ is an intermediate latent variable and so we are not free to choose a prior for it. Nevertheless, as far as inference on the right-hand-side of figure \ref{fig:ffbm}, we regard $p(b | X)$ as a pseudo-prior on $b$. We can show (appendix \ref{appdx:b|x}) that our choice of prior for $p(\theta)$ in equation \ref{eqn:theta-prior} leads to a uniform $p(b | X)$ in equation \ref{eqn:b-pseudo-prior}.
%
\begin{equation}
	p(b | X) = \int p(b | X, \theta) p(\theta) d\theta = B^{-N}
	\label{eqn:b-pseudo-prior}
\end{equation}
%
This is a uniform distribution that only depends on the number of blocks $B$. This is an enormously important simplification as evaluating $p(b | X)$ does not require knowing the exact value $X$ takes and can be performed without an expensive Monte-Carlo integration over the $\theta$-domain. \citet{Peixoto-Bayesian-Microcanonical} proposes careful choices for the additional microcanonical SBM parameters $\psi$ which we adopt. Peixoto's idea is to write the joint prior on $(b, e, k)$ as a product of conditionals $p(b, e, k) = p(b) p(e | b) p(k | e, b)= p(b) p(\psi | b)$. For our purposes we must insert a conditioning on $X$, to form our pseudo-prior for $b$ and $\psi$, to give equation \ref{eqn:joint-pseudo-prior}.
%
\begin{equation}
	p(b, \psi | X) = p(b | X) p(\psi | b, X) = p(b | X) p(\psi | b)
	\label{eqn:joint-pseudo-prior}
\end{equation}
%
Where the simplification is made apparent by noting $(\psi \indep X) | b$. We then borrow the priors proposed by \citet{Peixoto-Bayesian-Microcanonical} for $p(\psi | b)$ to complete our model. Please refer to appendix \ref{appdx:prior} for the exact form of $p(\psi | b)$. All that concerns us is that we have a form which is computable.
